Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-08 papers/termination/paper.tex
added captions and references to the figures with definitions

\end{framed}
\end{figure*}
 %This means that we will not be able to use the solver for the synthesis fragment to solve it.

@@

\begin{theorem}
\label{thm:ut}
 The loop $L(G, T)$ terminates from every start state iff formula {\bf [UT]} (Definition~\ref{def:UT}) is satisfiable.
\end{theorem}

@@

Since our program's state space is finite, a transition relation
induces an infinite execution iff some state is visited infinitely
often, or equivalently $ \exists x . T^+(x, x)$.
Deciding satisfiability of this formula directly would require a logic
that includes a transitive closure operator, $\bullet^+$.  Rather than
introduce such an operator, we will characterise non-termination
using the synthesis formula {\bf [ONT]} (Definition~\ref{def:ont})
encoding the existence of an \emph{(open) recurrence set}, i.e.~a nonempty 
set of states $N$ such that for each $s \in N$ there
exists a transition to some $s' \in N$ \cite{DBLP:conf/popl/GuptaHMRX08}.
%% Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterise non-termination
%% of a transition relation $T$ by the existence of an \emph{(open) recurrence
%% set}, i.e.~a nonempty set of states $N$ such that for each $s \in N$ there
%% exists a transition to some $s' \in N$.
%% The notion of open recurrence set is encoded by formula {\bf [ONT]} (Definition~\ref{def:ont}).

@@

If this formula is satisfiable, $N$ is an open recurrence set for $L$, which proves
$L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to the synthesis fragment
(it is an $\exists \forall \exists$ formula).  To eliminate the innermost existential quantifier,
we introduce a Skolem function $C$ that chooses the successor $x'$, which we then existentially quantify over.  This results in
formula {\bf [SNT]} (Definition~\ref{def:snt}).

@@

We note that if $T$ is deterministic, every open recurrence set is also a
closed recurrence set (since each state has at most one successor).  Thus,
the non-termination problem for deterministic transition systems is
equivalent to the satisfiability of formula {\bf [CNT]}.

@@

%% \begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
%% \label{def:multi-termination-formula}
%%  \begin{align*}
%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
%%                                                         & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
%% 							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
%%  \end{align*}
%% \end{definition}
 \end{framed}

\end{figure*}

@@

\caption{The non-termination formula for nested loops \label{fig:nonterm-nested}}
\end{figure*}

@@

\begin{theorem}
\label{thm:ct}
 The loop $L(G, T)$ terminates when started in any state satisfying $I(x)$ iff formula {\bf [CT]}
 (Defintion~\ref{def:ct}) is satisfiable.
\end{theorem}
% 
% \begin{proof}
%  Do this proof.
% \end{proof}

@@

\begin{proof}
%
Let $v_1, \ldots, v_k$ be the variables of the program $P$ under analysis,
and let each be $b$ bits wide.  Its state space $\mathcal{S}$ is then of
size $2^{bk}$.  A~ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$
exists iff $P$ terminates.  Without loss of generality,
$\mathcal{D}$ is a well-founded total order.  Since $R$ is injective, we have that $\|
\mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \|
\mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to
\mathcal{D'}$ with $ \| \mathcal{D'} = \| \mathcal{S} \|$ by just setting
$R' = R|_\mathcal{S}$, i.e.~$R'$ is just the restriction of $R$ to
$\mathcal{S}$.  Since $\mathcal{S}$ already comes equipped with a natural
well ordering we can also construct $R'' = \iota \circ R'$ where $\iota:
\mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from
$\mathcal{D'}$ to $\mathcal{S}$.  So assuming that $P$ terminates, there is
some ranking function $R''$ that is just a permutation of $\mathcal{S}$.  If
the number of variables $k > 1$ then in general the ranking function will be
lexicographic with dimension $\leq k$ and each co-ordinate of the output
being a single $b$-bit value.

@@

Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
$2^bk$ times where $k$ is the number of variables used by $L$ and
$b$ is the number of bits used to store each variable.
The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times l \times \| L\|$
with $l$ the length of the program to be synthesised, and $\| L \| $ the size of the program-under-analysis
computing $ L $.  So if
the CEGIS loop iterates $n$ times, the
runtime is bounded by $2 \times n \times NP(n \times l \times \| L \|)$, which is bounded
by $2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$.  Clearly the runtime is affected most strongly by $n$,
the total number of iterations of the CEGIS loop.

We will now show that the number of iterations of the CEGIS loop is a function
of the Kolmogorov complexity of the (non-)termination proof, and so
our runtime is asymptotically a function of the Kolmogorov complexity.
We argue that this gives our procedure various desirable qualities in practical
appliations.
We first recall the definition of the Kolmogorov complexity
of a function $f$:

@@

Let us consider the number of iterations of the CEGIS loop $n$ required for a specification
$\sigma$.  Since we enumerate candidate programs in order of length, we are always synthesising
programs with length no greater than $K(\sigma)$ (since when we enumerate the first correct program,
we will terminate).  So it is clear that the space of solutions we search over is the space
of functions computed by $\mathcal{L}$-programs of length no greater than $K(\sigma)$.  Let's
denote this set $\mathcal{L}(K(\sigma))$.  Clearly, $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$
since there are $O(2^{K(\sigma)}$ \emph{programs} of length $K(\sigma)$ and some functions
will be computed by more than one program.

Each iteration of the CEGIS loop distinguishes at least one incorrect function from the set of correct
functions, so we can see that the loop will iterate no more than $\| \mathcal{L}(K(\sigma)) \|$ times.
Therefore another bound on our runtime is:
$$O\left (2^{K(\sigma)} \times NP \left ( K(\sigma) \times \| L \| \right) \right) = O \left(2^{K(\sigma)} \right)$$
Which means that for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters
such as the size of the program under analysis.  To put this another way, under the assumption
that $K(\sigma)$ is small for a program's termination problem, we will find a termination proof
quite rapidly.

To give a flavour for the nature of the ``low Kolmogorov complexity'' assumption, 
we observe that linear functions have low Kolmogorov complexity (since a linear function can
be computed with at most 1 instruction and 1 multiplication for each program variable).
Conversely, the probability that a random program of length $l$ computes
a linear function is $O(2^{-l})$, since otherwise we'd need to have an exponential
number of programs computing the same function.

>>>>

\end{framed}
\caption{Formulae encoding the termination and non-termination of a single loop} \label{fig:single_loop}
\end{figure*}
 %This means that we will not be able to use the solver for the synthesis fragment to solve it.

@@

\begin{theorem}
\label{thm:ut}
 The loop $L(G, T)$ terminates from every start state iff formula {\bf [UT]} (Definition~\ref{def:UT}, Figure~\ref{fig:single_loop}) is satisfiable.
\end{theorem}

@@

Since our program's state space is finite, a transition relation
induces an infinite execution iff some state is visited infinitely
often, or equivalently $ \exists x . T^+(x, x)$.
Deciding satisfiability of this formula directly would require a logic
that includes a transitive closure operator, $\bullet^+$.  Rather than
introduce such an operator, we will characterise non-termination
using the synthesis formula {\bf [ONT]} (Definition~\ref{def:ont}, Figure~\ref{fig:single_loop})
encoding the existence of an \emph{(open) recurrence set}, i.e.~a nonempty 
set of states $N$ such that for each $s \in N$ there
exists a transition to some $s' \in N$ \cite{DBLP:conf/popl/GuptaHMRX08}.
%% Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterise non-termination
%% of a transition relation $T$ by the existence of an \emph{(open) recurrence
%% set}, i.e.~a nonempty set of states $N$ such that for each $s \in N$ there
%% exists a transition to some $s' \in N$.
%% The notion of open recurrence set is encoded by formula {\bf [ONT]} (Definition~\ref{def:ont}).

@@

If this formula is satisfiable, $N$ is an open recurrence set for $L$, which proves
$L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to the synthesis fragment
(it is an $\exists \forall \exists$ formula).  To eliminate the innermost existential quantifier,
we introduce a Skolem function $C$ that chooses the successor $x'$, which we then existentially quantify over.  This results in
formula {\bf [SNT]} (Definition~\ref{def:snt}, Figure~\ref{fig:single_loop}).

@@

We note that if $T$ is deterministic, every open recurrence set is also a
closed recurrence set (since each state has at most one successor).  Thus,
the non-termination problem for deterministic transition systems is
equivalent to the satisfiability of formula {\bf [CNT]} from Figure~\ref{fig:single_loop}.

@@

%% \begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
%% \label{def:multi-termination-formula}
%%  \begin{align*}
%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
%%                                                         & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
%% 							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
%%  \end{align*}
%% \end{definition}
 \end{framed}
\caption{Formula encoding conditional termination of a loop} \label{fig:conditional_termination}
\end{figure*}

@@

\caption{Formula encoding non-termination of nested loops \label{fig:nonterm-nested}}
\end{figure*}

@@

\begin{theorem}
\label{thm:ct}
 The loop $L(G, T)$ terminates when started in any state satisfying $I(x)$ iff formula {\bf [CT]}
 (Defintion~\ref{def:ct}, Figure~\ref{fig:conditional_termination}) is satisfiable.
\end{theorem}
% 
% \begin{proof}
%  Do this proof.
% \end{proof}

@@

\begin{proof}
%
Let $v_1, \ldots, v_k$ be the variables of the program $P$ under analysis,
and let each be $b$ bits wide.  Its state space $\mathcal{S}$ is then of
size $2^{bk}$.  A~ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$
exists iff $P$ terminates.  Since $R$ is injective, we have that $\|
\mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \|
\mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to
\mathcal{D'}$ with $ \| \mathcal{D'} = \| \mathcal{S} \|$ by just setting
$R' = R|_\mathcal{S}$, i.e.~$R'$ is just the restriction of $R$ to
$\mathcal{S}$.  Since $\mathcal{S}$ already comes equipped with a natural
well ordering we can also construct $R'' = \iota \circ R'$ where $\iota:
\mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from
$\mathcal{D'}$ to $\mathcal{S}$.  So assuming that $P$ terminates, there is
some ranking function $R''$ that is just a permutation of $\mathcal{S}$.  If
the number of variables $k > 1$ then in general the ranking function will be
lexicographic with dimension $\leq k$ and each co-ordinate of the output
being a single $b$-bit value.

@@

Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
$2^bk$ times where $k$ is the number of variables used by $L$ and
$b$ is the number of bits used to store each variable.
The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times \| L\|$
with $\| L \| $ the size of the program-under-analysis computing $\| L \|$.  So if 
the CEGIS loop iterates $n$ times, the
runtime is bounded by $2 \times n \times NP(n \times \| L \|)$, which is bounded
by $2^{bk + 1} NP(2^{bk} \| L \|)$.  Clearly the runtime is affected most strongly by $n$,
the total number of iterations of the CEGIS loop.

We will now show that our time complexity is asymptotically a function of
the Kolmogorov complexity of the (non-)termination proof, and argue
that this gives our procedure various desirable qualities in practical
appliations.
%
\begin{theorem}
 {\sc Kalashnikov} has time complexity $NP^{NP}(K(\sigma))$.
\end{theorem}

\begin{theorem}
 {\sc Headshot} uses a single call to {\sc Kalashnikov} and uses an encoding with $O(1)$
 overhead.  It therefore has time complexity $NP^{NP}(K(\tau(P))$ where $\tau(P)$ is the
 termination specification for program $P$.
\end{theorem}

In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
This procedure has the property that it will find the shortest program meeting the
specification, and as a consequence the runtime of the program synthesiser is dominated
by the size of this shortest program.  To frame the discussion of the runtime of our
termination proving procedure, we first recall the definition of the Kolmogorov complexity
of a function $f$:

@@

We will later show that the time complexity of our synthesis procedure is $NP^{NP}(K(\sigma))$,
which means that for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters
such as the size of the program under analysis.  To put this another way, under the assumption
that $K(\sigma)$ is small for a program's termination problem, we will find a ranking function
quite rapidly.  We will now investigate the properties of functions with low Kolmogorov complexity,
and show that the assumption of a low-Kolmogorov ranking function is much weaker than the assumption
of a linear ranking function.

\begin{theorem}
 Linear functions have low Kolmogorov complexity.
\end{theorem}

\begin{proof}
 The function $f: X \to Y$ can be computed with a program consisting of
 $2 \cdot \dim(X) \cdot \dim(Y) - \dim(Y)$ instructions (1 multiplication and 1 addition for
 each cell in the matrix representing $f$).  Therefore,
 $K(f) \leq 2 \cdot \dim(X) \cdot \dim(Y) - \dim(Y)$.
\end{proof}

\begin{theorem}
Assuming a reasonable program encoding, the probability that a random program of length $l$ computes
a linear function is $O(2^{-l})$.
\end{theorem}

\begin{proof}
 (Do this proof properly).

 Otherwise, we'd need to have an exponential number of programs computing the same function.
 That would be an unreasonable program encoding.
\end{proof}
